Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2020-10-12 paper/paper.Rmd
update on methods

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds beyond the capacity of wind or other abiotic processes. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersa events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: 
  word_document:
    reference_docx: paper_ref.docx
bibliography: mybibfile.bib
---

@@

Seed dispersal mutualisms may be a particularly attractive system to study the impacts of incorporating individual variation to understanding the importance of such variation to dispersal outcomes. Seed dispersal is one of the most critical stages in plant life history – it results in a spatial pattern of seed deposition which acts as a template that will define the distribution of plants and, consequently, community structure [@howe1982ecology; @nathan_spatial_2000; @russo_incorporating_2006]. In seed dispersal mutualisms, animals provide the dispersal services while rewarding from fruit resources they consume. Plants produce fleshy fruits to attract and reward mutualist animals and rely on the behavior of the disperser to transport their propagules [@nathan_spatial_2000]. The spatial patterns of seed deposited by disperser vectors will determine the probability of seed survival and recruitment, acting upon post-dispersal processes such as density-dependent survival and colonization [@howe2004seed]. Therefore, at a local scale, seed dispersers will determine whether plant species scape from density-dependence processes [@comita2014testing], whereas at a larger scale, they can determine how quickly plants can cope with habitat and climate shifts [@ibanez2006predicting; @russo_incorporating_2006].

From the plant perspective, not all seed disperser species provide with the same seed dispersal service, and often relatively few disperser species can have disproportionate effects on seed dispersal [@schupp1993quantity; @jordano2007differential; @schupp_seed_2010; @loayza2014seed]. Frugivores might differ on their selection of plants/fruits for consumption, in their treatment of seed while ingesting and manipulating the fruits, in their processing of seeds if ingested and in the distance where they will discard the seeds [@schupp1993quantity; @schupp_seed_2010; @cortes_integrating_2013]. Each one of these stages of the seed dispersal process will have strong consequences on the spatial seed shadow provided by each seed dispersal vector [@jordano2007differential; @carlo2014directness]. Besides inter-specific differences among dispersal vectors, there is a recent recognition of the potential role of differences among individuals in the resulting seed dispersal they provide [@bolnick_ecology_2003; @gonzalez-varo_labile_2016]. Including individual variation in traits (morphological, behavioral and physiological) in seed dispersal studies, can provide new insights into the relative role of different dispersal vectors on seed deposition and ultimately, vegetation structure [@gonzalez-varo_labile_2016]. In a recent review, Zwolak (2017) summarized the types of intraspecific variation in seed-dispersing animals that might affect the resulting seed dispersal service individuals provide, highlighting differences in sex, size or age (ontogenetic shifts), individual specialization and behavioral syndromes as the most important variation types. Incorporating these intrinsic traits into seed dispersal models may lead to more mechanistic understanding of seed dispersal, allow to build more accurate predictive frameworks of dispersal outcomes [@russo_incorporating_2006] and to identify links between characteristics of dispersal agents and the seed dispersal they provide [@zwolak_how_2018].

Differences in behavior among individuals (i.e. behavioral syndromes sensu [@sih2004behavioral]) can have remarkable consequences on ecological and evolutionary processes [@wolf_animal_2012]. For vertebrate dispersed plants, animal behavior will inﬂuence seed shadows and few species, or even individuals, might provide unique, high value dispersal services by moving farther distances from parental plants [@schupp_seed_2010]. Long distance dispersal (LDD) events are crucial to reaching suitable sites to germinate and establish and to colonize new habitats [@nathan2006long]. Furthermore, rare events (often overlooked) are critical for dispersal and might have a large effect on resulting plant demography [@loayza2014seed]. Incorporating animal behavior and the individual distances they carry seeds into seed dispersal models might result in better predictions of seed shadow and better inform about the resulting patterns of seed dispersal.

The effect of frugivore movement on seed dispersal has been studied and we know that there exists an important link between them [@will_mechanistic_2008; @russo_incorporating_2006; @pegman_exploring_2017]. Animal movement patterns determine the extent to which seeds get distributed, and this in turn creates a range distribution for plants [@levey_modelling_2008]. Plants distribution and ranges are largely determined by the seeds that get dispersed the farthest from their parent. Animal movement is known to provide these long-distance seeds more than other dispersal sources such as wind or water[@jordano_what_2017]. Frugivore in particular is associated with longer dispersal, partly influenced by the gut retention time. We wanted to understand how individual variation in animal movement can influence the frequency and extent of long-distance seed dispersal events. We analyzed aracari telemetry data at the population level and at an individual level, to explore how these could influence seed dispersal. In addition to this, we tested several models to understand how the choice of probability distribution associated to the movement kernel can limit or overestimate seed dispersal.

# Methods
## Seed dispersal simulation
the goal of this manuscript is to evaluate the effects of variation in animal movement and how these translate to seed dispersal distances. We developed an individual-based simulation in order to track single individuals or family groups in various tracking sessions and see how they disperse seeds.

>>>>

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: 
  word_document:
    reference_docx: paper_ref.docx
bibliography: mybibfile.bib
---

@@

Seed dispersal mutualisms may be a particularly attractive system to study the impacts of incorporating individual variation to understanding the importance of such variation to dispersal outcomes. Seed dispersal is one of the most critical stages in plant life history – it results in a spatial pattern of seed deposition which acts as a template that will define the distribution of plants and, consequently, community structure [@howe1982ecology; @nathan_spatial_2000; @russo_incorporating_2006]. In seed dispersal mutualisms, animals provide the dispersal services while rewarding from fruit resources they consume. Plants produce fleshy fruits to attract and reward mutualist animals and rely on the behavior of the disperser to transport their propagules [@nathan_spatial_2000]. The spatial patterns of seed deposited by disperser vectors will determine the probability of seed survival and recruitment, acting upon post-dispersal processes such as density-dependent survival and colonization [@howe2004seed]. Therefore, at a local scale, seed dispersers will determine whether plant species escape from density-dependence processes [@comita2014testing], whereas at a larger scale, they can determine how quickly plants can cope with habitat and climate shifts [@ibanez2006predicting; @russo_incorporating_2006].

From the plant perspective, not all seed dispersers species provide with the same seed dispersal service, and often relatively few disperser species can have disproportionate effects on seed dispersal [@schupp1993quantity; @jordano2007differential; @schupp_seed_2010; @loayza2014seed]. Frugivores might differ on their selection of plants/fruits for consumption, in their treatment of seed while ingesting and manipulating the fruits, in their processing of seeds if ingested and in the distance where they will discard the seeds [@schupp1993quantity; @schupp_seed_2010; @cortes_integrating_2013]. Each one of these stages of the seed dispersal process will have strong consequences on the spatial seed shadow provided by each seed dispersal vector [@jordano2007differential; @carlo2014directness]. Besides inter-specific differences among dispersal vectors, there is a recent recognition of the potential role of differences among individuals in the resulting seed dispersal they provide [@bolnick_ecology_2003; @gonzalez-varo_labile_2016]. Including individual variation in traits (morphological, behavioral and physiological) in seed dispersal studies, can provide new insights into the relative role of different dispersal vectors on seed deposition and ultimately, vegetation structure [@gonzalez-varo_labile_2016]. In a recent review, Zwolak [@zwolak_how_2018] summarized the types of intraspecific variation in seed-dispersing animals that might affect the resulting seed dispersal service individuals provide, highlighting differences in sex, size or age (ontogenetic shifts), individual specialization and behavioral syndromes as the most important variation types. Incorporating these intrinsic traits into seed dispersal models may lead to more mechanistic understanding of seed dispersal, allow to build more accurate predictive frameworks of dispersal outcomes [@russo_incorporating_2006] and to identify links between characteristics of dispersal agents and the seed dispersal they provide [@zwolak_how_2018].

Differences in behavior among individuals (i.e. behavioral syndromes sensu [@sih2004behavioral]) can have remarkable consequences on ecological and evolutionary processes [@wolf_animal_2012]. For vertebrate dispersed plants, animal behavior will inﬂuence seed shadows and few species, or even individuals, might provide unique, high value dispersal services by moving farther distances from parental plants [@schupp_seed_2010]. Long distance dispersal (LDD) events are crucial to reaching suitable sites to germinate and establish and to colonize new habitats [@nathan2006long]. Furthermore, rare events (often overlooked) are critical for dispersal and might have a large effect on resulting plant demography [@loayza2014seed]. Incorporating animal behavior and the individual distances they carry seeds into seed dispersal models might result in better predictions of seed shadow and better inform about the resulting patterns of seed dispersal. To understand the consequences of variation in animal movement, we compare seed dispersal estimates from simulation models that incorporate seed retention time and individual, family, or population level animal movement.

# Methods
## Seed dispersal simulation

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our simulation time steps were based on one minute intervals, and each simulation started at the origin or source tree, where each individual bird was given five seeds and was allowed to move in the landscape until the simulation ended once all seeds were dropped. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution. Commonly used probability distributions for sampling the animal movement step size included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size, with distributions such as the lognormal approaching a levy type walk (not sure this is all correct or makes sense).
